Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2021-07-27 analysis/paper/shceof.tex
Muchos cambios

EOFs are more flexible than Fourier descomposition modes in the sense that they can characterise \textcolor{red}{Esto hay que exresarlo más claramente}.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
Besides, a single EOF can also represent a mixture of two or more physical modes.
The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
\citet{irving2016} characterises the Pacific--South American Pattern in a Fourier framework.
Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
On this new projection, they could identify the PSA using Fourier decomposition.

@@

\textcolor{red}{The data and methods are described in Section 2 :P ...}

\hypertarget{data-and-methods}{%
\section{Data and Methods}\label{data-and-methods}}

@@

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

@@

\hypertarget{ceof-spatial-patterns}{%
\subsection{cEOF spatial patterns}\label{ceof-spatial-patterns}}

@@

\hypertarget{ceofs-regressions}{%
\subsection{cEOFs Regressions}\label{ceofs-regressions}}

@@

\hypertarget{ceofs-relationship-with-other-variables}{%
\subsection{cEOFs relationship with other variables}\label{ceofs-relationship-with-other-variables}}

\textcolor{red}{Quedo medio rara la divsión con los títulos. Me parece que convendrá discutir primero la relación con otros modos de circulación conocidos (SAM, Ozono) y después con las otras variables(Temp, TOC, SST, precip)}

@@

\caption{\label{tab:psa-eof2}Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep[following][]{mo2001} for the 1979 -- 2019 period. 95\% confidence intervals in parenthesis.
Background colour indicates the value of correlation.}
\centering
\begin{tabular}[t]{l>{}l>{}l}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{cEOF2} \\
\cmidrule(l{3pt}r{3pt}){2-3}
PC & Real & Imaginary\\
\midrule
PSA1 & \cellcolor[HTML]{C3BBDE}{\textcolor{black}{-0.32 (CI: -0.57 -- -0.01)}} & \cellcolor[HTML]{6058A9}{\textcolor{white}{-0.83 (CI: -0.9 -- -0.69)}}\\
PSA2 & \cellcolor[HTML]{645BAB}{\textcolor{white}{-0.81 (CI: -0.9 -- -0.67)}} & \cellcolor[HTML]{F9F1F0}{\textcolor{black}{0.059 (CI: -0.25 -- 0.36)}}\\
\bottomrule
\end{tabular}
\end{table}

The result obtaneid here is similar to the one obtained by \citet{irving2016}.
From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

\hypertarget{is-the-asymmetric-sam-real}{%
\subsubsection{Is the Asymmetric SAM real?}\label{is-the-asymmetric-sam-real}}

\begin{figure}
\centering
\includegraphics{../figures/fake-eof-1.pdf}
\caption{\label{fig:fake-eof}Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa. Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis. Units are arbitrary.}
\end{figure}

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated annular pattern, we create a synthetic dataset of 41 fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric annylar pattern with random sign and value sampled with replacement from the real SAM index.
To this synthetic dataset we then apply traditional EOF.
By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure~\ref{fig:fake-eof} row a).
While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so traditional EOF analysis needs two EOFs to represented it correctly.
Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

The three EOFs of the synthetic dataset are shown in Figure~\ref{fig:fake-eof} row b.
The annular pattern is captured mainly by the first EOF while the cEOF2 is explained by the second and third.
However, even though the annular pattern used to construct these data is perfectly zonally symmetric, the leading EOF is not.
The EOF decomposition mixes the true annular pattern with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern \citep{fogt2020}.

\textcolor{red}{Todo esto está mal.}

Rotating makes it even worse.

The reason this happens is that the spatial correlation between the annular pattern and the wave patterns is not zero; i.e.~they are not orthogonal. And since EOFs

In this particular example, it is 0.022 (CI: -0.29 -- 0.33) with the Real part and 0.016 (CI: -0.29 -- 0.32) with the Imaginary part.
Of note, the three EOFs are well separated according to North's rule of thumb \citep{north1982}.

This suggests that the zonal asymmetries of the SON SAM could very well be mostly a statistical artefact.
Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO \citep{campitelli2021}, this potential issue could affect the interpretability of many results involving the SAM.
Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.

\hypertarget{tropical-sources}{%
\subsection{Tropical sources}\label{tropical-sources}}

@@

The connections between cEOFs and Tropical sources is also assesed.
Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2. The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2). This pattern is almost canonically positive ENSO \citep{bamston1997} and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.87).
Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2). This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000}.

@@

Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
The correlation between the Real cEOF2 and ENSO is also not significant (-0.0042 (CI: -0.31 -- 0.3)).
This lack of tropical signal suggests different natures of the different phases of the Real cEOF2 wave train, which has the maximum amplitudes in the surroundings of southern South America.

\begin{figure}
\centering
\includegraphics{../figures/enso-phase-1.pdf}
\caption{\label{fig:enso-phase} ONI plotted against cEOF2 phase for th 1979 -- 2019 period. Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively. Black line is the fit ONI \textasciitilde{} sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
\end{figure}

@@

The influence of cEOFs variability in continental rainfall of extra-tropical Southern Hemisphere is also explored.
We focus on South America and Oceania sectors, because there are not relevant signals in Southern Africa (not shown).

@@

The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and widespread anomalies. The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia. Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021} and ENSO \citep{cai2011}.

\begin{figure}
\centering
\includegraphics{../figures/australia-pp-phase-1.pdf}
\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitation anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b).
Black line is the fit precipitation \textasciitilde{} sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
\end{figure}

@@

Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each cEOF in South America. cEOF1 is associated with slight decreases \textcolor{red}{Revisa como estás expresando toda esta parte (y en el resto del texto), recorda que son regresiones así que no corresponde decir postive/negative sino haces alguna referencia a la variable con las que los estás regresionando.} in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column a). The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2. The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
On the other hand, the Real cEOF2 is associated with negative precipitation anomalies in a smaller area of SESA.
Resembling the relationship between ONI and the phase of cEOF2 (Fig.~\ref{fig:enso-phase}), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
These variable could be related wth a signficant sinusoidal fit that has a coefficient of determination of~0.27 (p-value~= 0.0022).

\textcolor{red}{La formula en la figura trata de escribirla en una forma más precisa "precipitación"/"phase" son muy generales, y también teniendo en cuenta las unidades. También incluilo en el caption.}

@@

THe second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
Essentially, it is an alternate representation of the PSA1 and PSA2 patterns \citep{mo2001}.
We show that the Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
This cEOF prefers a phase aligned with the Imaginary cEOF2.
This is because the when ENSO is active, it forces the cEOF2 to be in the Imaginary phase.
When ENSO is neutral, the cEOF2 is still active, but with no particular phase.
This mirrors the results of \citet{cai2002}, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.

>>>>

EOFs are more flexible than Fourier descomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.
The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
\citet{irving2016} characterises the Pacific--South American Pattern in a Fourier framework.
Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
On this new projection, they could identify the PSA using Fourier decomposition.

@@

In Section \ref{methods} we describe the methods.
In Section \ref{temporal} we describe the temporal variability and vertical coherence of the indices.
In Section \ref{spatial}, we analyse the spatial patterns of each complex EOF.
In Section \ref{regressions} we study the spatial regressions with geopotential height and temperatuer.
In Section \ref{other-variables}, we analyise the relationship between cEOF1 and Ozone, and between EOF2 and the Southern Annular Mode and the PSA.
In Section \ref{tropical} we look for tropical forcings that explain the variability of each cEOF.
Finally, in Section \ref{precipitation} we show the relationship between these modes of variabiltiy and precipitation in South America and Oceania.

\hypertarget{methods}{%
\section{Data and Methods}\label{methods}}

@@

\hypertarget{methods-1}{%
\subsection{Methods}\label{methods-1}}

@@

\hypertarget{spatial}{%
\subsection{cEOF spatial patterns}\label{spatial}}

@@

\hypertarget{regressions}{%
\subsection{cEOFs Regressions}\label{regressions}}

@@

\hypertarget{other-variables}{%
\subsection{cEOFs relationship with other variables}\label{other-variables}}

\caption{\label{tab:psa-eof2}Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep[following][]{mo2001} for the 1979 -- 2019 period. 95\% confidence intervals in parenthesis.
Background colour indicates the value of correlation.}
\centering
\begin{tabular}[t]{l>{}l>{}l}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{cEOF2} \\
\cmidrule(l{3pt}r{3pt}){2-3}
PC & Real & Imaginary\\
\midrule
PSA1 & \cellcolor[HTML]{C3BBDE}{\textcolor{black}{-0.32 (CI: -0.57 -- -0.015)}} & \cellcolor[HTML]{6259AA}{\textcolor{white}{-0.82 (CI: -0.9 -- -0.69)}}\\
PSA2 & \cellcolor[HTML]{9E4E47}{\textcolor{white}{0.81 (CI: 0.67 -- 0.9)}} & \cellcolor[HTML]{F3F2F8}{\textcolor{black}{-0.062 (CI: -0.36 -- 0.25)}}\\
\bottomrule
\end{tabular}
\end{table}

The result obtained here is similar to the one obtained by \citet{irving2016}.
From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

@@

\hypertarget{tropical}{%
\subsection{Tropical sources}\label{tropical}}

@@

The connections between cEOFs and Tropical sources is also assessed.
Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2. The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2). This pattern is almost canonically positive ENSO \citep{bamston1997} and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.87).
Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2). This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000}.

@@

Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
The correlation between the Real cEOF2 and ENSO is also not significant (0.00025 (CI: -0.31 -- 0.31)).
This lack of tropical signal suggests different natures of the different phases of the Real cEOF2 wave train, which has the maximum amplitudes in the surroundings of southern South America.

\begin{figure}
\centering
\includegraphics{../figures/enso-phase-1.pdf}
\caption{\label{fig:enso-phase}ONI plotted against cEOF2 phase for th 1979 -- 2019 period. Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively. Black line is the fit ONI \textasciitilde{} sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
\end{figure}

@@

The influence of cEOFs variability in continental rainfall of extra-tropical Southern Hemisphere is also explored.
We focus on South America and Oceania sectors because there are no relevant signals in southern Africa (not shown).

@@

The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and widespread anomalies. The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia. Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021} and ENSO \citep{cai2011}.

\begin{figure}
\centering
\includegraphics{../figures/australia-pp-phase-1.pdf}
\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitation anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b) in \(\frac{\mathrm{mm}}{\mathrm{day}}\).
Black line is the fit precipitation \textasciitilde{} sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
\end{figure}

@@

Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each cEOF in South America. cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column a). The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2. The positive regression coefficients on Southeastern South America (SESA) and Central Chile, and negative ones over South Atlantic Convergence Zone is a well-known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
On the other hand, the Real cEOF2 is associated with negative precipitation anomalies in a smaller area of SESA.
Resembling the relationship between ONI and the phase of cEOF2 (Fig.~\ref{fig:enso-phase}), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
These variables could be related wth a significant sinusoidal fit that has a coefficient of determination of~0.27 (p-value~= 0.0022).

@@

The second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
Essentially, it is an alternate representation of the PSA1 and PSA2 patterns \citep{mo2001}.
We show that the Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
This cEOF prefers a phase aligned with the Imaginary cEOF2.
This is because the when ENSO is active, it forces the cEOF2 to be in the Imaginary phase.
When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
This mirrors the results of \citet{cai2002}, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal. The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except frot the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
